---
title: "Gender Bias in Graduate Admissions"
author: "Adelson Aguasvivas"
date: "March 7th, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

data(UCBAdmissions)
library(broom)
library(dplyr)
library(ggplot2)
library(scales)
library(forcats)

ucb_tidy <- tidy(UCBAdmissions)

```

# Goal

- To explore whether there was gender bias or discrimination in Graduatr
Admissions at UC Berkley

```{r acceptance-rate, echo=FALSE}

ucb_tidy_aggregated <- ucb_tidy %>% 
  group_by(Admit, Gender) %>% 
  summarize(n = sum(n)) %>% 
  ungroup() %>% 
  group_by(Gender) %>% 
  mutate(prop = n / sum(n)) %>% 
  filter(Admit == "Admitted")

gg_bar <- ucb_tidy_aggregated %>% 
    ggplot(aes(x = Gender, y = prop, fill = Gender)) +
    geom_col() +
    geom_text(aes(label = percent(prop)), vjust = -1) +
    labs(title = "Acceptance rate of male and female applicants",
         subtitle = "University of California, Berkeley (1973)",
         y = "Acceptance rate") +
    scale_y_continuous(labels = percent, limits = c(0, 0.5)) +
    guides(fill = FALSE)

gg_bar
```

```{r faceted-aceptance, echo=FALSE}

ucb_by_dept <- ucb_tidy %>% 
    group_by(Gender, Dept) %>% 
    mutate(prop = n / sum(n)) %>% 
    filter(Admit == "Admitted")

gg_bar_faceted <- ucb_by_dept %>% 
  ggplot(aes(Gender, prop, fill = Gender)) +
  geom_col() +
  geom_text(aes(label = percent(prop)), vjust = -1) +
  labs(title = "Acceptance rate of male and female applicants",
       subtitle = "University of California, Berkeley (1973)",
       y = "Acceptance rate") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  facet_wrap(~ Dept) +
  guides(fill = FALSE)

gg_bar_faceted

```

```{r bias-discrimination, echo=FALSE}

multiply_rows <- function(column, n) {
  rep(column, n)
}

ucb_full <- data.frame(Admit = multiply_rows(ucb_tidy$Admit, ucb_tidy$n),
                      Gender = multiply_rows(ucb_tidy$Gender, ucb_tidy$n),
                      Dept = multiply_rows(ucb_tidy$Dept, ucb_tidy$n))

# Reverse the coding of the Admit variable
ucb_full$Admit <- fct_relevel(ucb_full$Admit,"Rejected", "Admitted")

# Run the regression
glm_gender <- glm(Admit ~ Gender, data = ucb_full, family = "binomial")

# Run the regression, including Dept as an explanatory variable
glm_genderdept <- glm(Admit ~ Gender + Dept, 
                      data = ucb_full, 
                      family = "binomial")

# Filter for Department A
dept_a <- ucb_full %>%
  filter(Dept == "A")

# Run the regression
glm_gender_depta <- glm(Admit ~ Gender, 
                      data = dept_a, 
                      family = "binomial")

glm_gender

glm_genderdept

glm_gender_depta

# Define bias
bias <- "a pattern of association between a 
particular decision and a particular sex of applicant, 
of sufficient strength to make us confident that it is 
unlikely to be the result of chance alone"

# Define discrimination
discrimination <- "the exercise of decision
influenced by the sex of the applicant
when that is immaterial to the qualifications for entry"


```

# Conclusion

- Bias is defined as a pattern of association between a 
particular decision and a particular sex of applicant, 
of sufficient strength to make us confident that it is 
unlikely to be the result of chance alone

- Discrimination is defined as the exercise of decision
influenced by the sex of the applicant
when that is immaterial to the qualifications for entry

- Bias is not the same as Discrimination